---
author: robjhyndman
comments: true
date: 2014-02-21 10:36:14+00:00
mathjax: true
link: https://robjhyndman.com/hyndsight/forecasting-within-limits/
slug: forecasting-within-limits
title: Forecasting within limits
wordpress_id: 2492
categories:
- forecasting
- R
- statistics
---

It is common to want forecasts to be positive, or to require them to be within some specified range $[a,b]$. Both of these situations are relatively easy to handle using transformations.<!-- more -->



### Positive forecasts



To impose a positivity constraint, simply work on the log scale. With the forecast package in R, this can be handled by specifying the Box-Cox parameter $\lambda=0$. For example, consider the real price of a dozen eggs (1900-1993; in cents):

```{r, message=FALSE, cache=TRUE}
    library(fpp)
    fit <- ets(eggs, lambda=0)
    plot(forecast(fit, h=50))
```    


### Forecasts constrained to an interval



To see how to handle data constrained to an interval, imagine that the egg prices were constrained to lie within $a=50$ and $b=400$. Then we can transform the data using a scaled logit transform which maps $(a,b)$ to the whole real line:
$$
y = \log\left(\frac{x-a}{b-x}\right)
$$
where $x$ is on the original scale and $y$ is the transformed data.


```{r, cache=TRUE}
    # Bounds
    a <- 50
    b <- 400
    # Transform data
    y <- log((eggs-a)/(b-eggs))
    fit <- ets(y)
    fc <- forecast(fit, h=50)
    # Back-transform forecasts
    fc$mean <- (b-a)*exp(fc$mean)/(1+exp(fc$mean)) + a
    fc$lower <- (b-a)*exp(fc$lower)/(1+exp(fc$lower)) + a
    fc$upper <- (b-a)*exp(fc$upper)/(1+exp(fc$upper)) + a
    fc$x <- eggs
    # Plot result on original scale
    plot(fc)
```    


The prediction intervals from these transformations have the same coverage probability as on the transformed scale, because quantiles are preserved under monotonically increasing transformations.
