---
author: robjhyndman
comments: true
date: 2016-06-01 07:22:52+00:00
slug: forecast7-part-2
title: Forecast v7 (part 2)
categories:
- forecasting
- graphics
- R
- time series
---

As mentioned in [my previous post on the forecast package v7](https://robjhyndman.com/hyndsight/forecast7-ggplot2/), the most visible feature was the introduction of [ggplot2 graphics](http://ggplot2.org/). This post briefly summarizes the remaining new features of forecast v7.

```{r, message=FALSE}
library(forecast)
library(ggplot2)
```

### tslm rewritten

The `tslm` function is designed to fit linear models to time series data. It is intended to approximately mimic `lm` (and  calls `lm` to do the estimation), but to package the output to remember the `ts` attributes. It also handles some predictor variables automatically, notably `trend` and `season`. The re-write means that `tslm` now handles functions as predictors, including `fourier`.

```{r, cache=TRUE}
deaths.lm  <- tslm(mdeaths ~ trend + fourier(mdeaths,3))
mdeaths.fcast <- forecast(deaths.lm,
    data.frame(fourier(mdeaths,3,36)))
autoplot(mdeaths.fcast)
```

Note that `fourier` now takes 3 arguments. The first is the series, which is only used to grab the seasonal period and the `tsp` attribute. The second argument `K` is the number of Fourier harmonics to compute. If the third argument `h` is `NULL` (the default), the function returns Fourier terms for the times of the historical observations. But if `h` is a positive integer, the function returns Fourier terms for the next `h` time periods after the end of the historical data.

The `lm` function has long allowed a matrix to be passed and independent linear models fitted to each column. The new `tslm` function also allows this now.

### Bias adjustment for Box-Cox transformations

Almost all modelling and forecasting functions in the package allow Box-Cox transformations to be applied before the model is fitted, and for the forecasts to be back transformed. This will give median forecasts on the original scale, as [I've explained before](https://robjhyndman.com/hyndsight/backtransforming/).

There is now an option to adjust the forecasts so they are means rather than medians, but setting `biasadj=TRUE` whenever the forecasts are computed. I will probably make this the default in some future version, but for now the default is `biasadj=FALSE` so the forecasts are actually medians.

```{r, message=FALSE, cache=TRUE}
library(fpp, quietly=TRUE)
fit <- ets(eggs, model="AAN", lambda=0)
fc1 <- forecast(fit, biasadj=TRUE, h=20, level=95)
fc2 <- forecast(fit, biasadj=FALSE, h=20)
cols <- c("Mean"="#0000ee","Median"="#ee0000")
autoplot(fc1) + ylab("Price") + xlab("Year") +
  autolayer(fc2, PI=FALSE, series="Median") +
  autolayer(fc1, PI=FALSE, series="Mean") +
  guides(fill=FALSE) +
  scale_colour_manual(name="Forecasts",values=cols)
```

### A new Ccf function

Cross-correlations can now be computed using `Ccf`, mimicing `ccf` except that the axes are more informative.

The `Acf` function now handles multivariate time series, with cross-correlation functions computed as well as the ACFs of each series.

### Covariates in neural net AR models

The `nnetar` function allows neural networks to be applied to time series data by building a nonlinear autoregressive model. A new feature allows additional inputs to be included in the model.

### Better subsetting of time series

`subset.ts` allows quite sophisticated subsetting of a time series. For example

```{r, cache=TRUE}
plot(subset(gas,month="November"))
subset(woolyrnq,quarter=3)
```

This is now substantially more robust than it used to be.

### What's next?

The next major release will probably be around the end of 2016. On the to-do list are:

  * **In-sample multi-step fitted values**. Currently `fitted` returns in-sample one-step forecasts. A new argument to `fitted` will allow multi-step forecasts of the training data.

  * **Applying fitted models to new data sets**. A related issue is to take an estimated model and apply it to some new data without re-estimating parameters. This is already possible with `Arima` and `ets` models. It will be extended to many more model types.

  * **Better choice of seasonal differencing**. Currently `auto.arima` does a pretty good job at finding the orders of a model, and the number of first-differences required, but it does not handle seasonal differences well. It often selects 0 differences, when I think it should select 1 difference. So I tend to over-ride the automatic choice with `auto.arima(x, D=1)`. I will attempt to find some better tests of seasonal unit roots than those that are currently implemented.

  * **Prediction intervals for NNAR forecasts**. The forecasts obtained using a NNAR model (via the `nnetar` function) do not have prediction intervals because there is no underlying stochastic model on which to base them. However, there are ways of computing the uncertainty using simulation, and I hope to implement something like that for the next version.

