---
date: 2018-12-18
title: Why doesn't auto.arima() return the model with the lowest AICc value?
slug: badarima
---

This question seems to come up frequently on [crossvalidated.com](http://crossvalidated.com) or in my inbox.

> I have this time series, however it yields different results when I use the `auto.arima` and `Arima` functions.

```{r badarima}
library(forecast)
xd <- ts(c(23786, 25955, 54373, 21561, 14552, 13284, 12714, 11821, 15445, 21307, 17228, 20007, 23065, 32811, 43147, 15127, 13497, 12224, 11412, 11888, 14210,18978, 15782, 17216, 16417, 22861, 42616, 17057,  9741, 10503,  7170, 10686,  9762, 15773, 15280, 13212, 14784, 26104, 29947), frequency = 12, start=c(2014,1), end=c(2017,3))

fit1 <- auto.arima(xd, trace=TRUE, allowmean=FALSE, allowdrift = FALSE, max.p=1, max.q=0, max.P=1, max.Q=0, xreg=seq_along(xd))

(fit2 <- Arima(xd, order=c(1,0,0), seasonal=c(1,0,0), xreg=seq_along(xd)))
```

> `auto.arima()` shows an AICc value of `Inf` for an ARIMA(1,0,0)(1,0,0) model, while the same model has a finite value using `Arima()`.

The issue here is to do with the checks carried out by `auto.arima()` in an effort to return a good model. The `auto.arima()` function does not simply find the model with the lowest AICc value. It also carries out several checks to ensure the model is numerically well-behaved.

While the `Arima()` function will never return a model with roots inside the unit circle, the `auto.arima()` function is even stricter and will not select a model with roots close to the unit circle either. The ARIMA(1,0,0)(1,0,0)$_{12}$ model fitted above has roots almost on the unit circle. This is easily seen by plotting them.

```{r plotroots}
Arima(xd, order=c(1,0,0), seasonal=c(1,0,0), xreg=seq_along(xd)) %>%
  autoplot()
```

In fact, there are 12 roots with absolute value `r sprintf("%4.3f", abs(forecast:::arroots(fit2)$roots)[1])`, just outside the unit circle (so the inverse roots that are plotted are just inside the circle). Consequently, this model is rejected by `auto.arima()` because the forecasts will be numerically unstable, and the AICc value is set to `Inf` to prevent it being selected.

In general, unless you know exactly what you're doing, it is better to leave `auto.arima()` to select a model for you. In this case, it comes up with the following model which should forecast well.

```{r bestmodel}
(fit <- auto.arima(xd))
fit %>% forecast() %>% autoplot()
```