---
author: robjhyndman
comments: true
date: 2020-02-07
title: "Electricity demand data in tsibble format"
slug: electrictsibbles
mathjax: true
categories:
  - time series
  - statistics
  - R
  - tidyverts
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
options(digits = 3, width = 75)
```

The `tsibbledata` packages contains the `vic_elec` data set, containing half-hourly electricity demand for the state of Victoria, along with corresponding temperatures from the capital city, Melbourne. These data cover the period 2012-2014.

Other similar data sets are also available, and these may be of interest to researchers in the area.

For people new to tsibbles, please read my [introductory post](https://robjhyndman.com/hyndsight/tsibbles/).

&nbsp;

## Australian state-level demand

The rawdata for other states are also stored in the [`tsibbledata` github repository](https://github.com/tidyverts/tsibbledata/) (under the data-raw folder), but these are not included in the package to satisfy CRAN space constraints. However, anyone can still load and use the data with the following code.

```{r loadpackages, message=FALSE}
library(tidyverse)
library(lubridate)
library(tsibble)
```

```{r ausdata, message=FALSE}
repo <- "https://raw.githubusercontent.com/tidyverts/tsibbledata/master/data-raw/vic_elec/"
states <- c("NSW","QLD","SA","TAS","VIC")
dirs <- paste0(repo, states, "2015")

# Read holidays data
holidays <- paste0(dirs,"/holidays.txt") %>%
  as.list() %>%
  map_dfr(read_csv, col_names=FALSE, .id="State") %>%
  transmute(
    State = states[as.numeric(State)],
    Date = dmy(X1), 
    Holiday = TRUE
  )
# Read temperature data
temperatures <- paste0(dirs,"/temperature.csv") %>%
  as.list() %>%
  map_dfr(read_csv, .id = "State") %>%
  mutate(
    State = states[as.numeric(State)],
    Date = as_date(Date, origin = ymd("1899-12-30"))
  )
# Read demand data
demands <- paste0(dirs,"/demand.csv") %>%
  as.list() %>%
  map_dfr(read_csv, .id = "State") %>%
  mutate(
    State = states[as.numeric(State)],
    Date = as_date(Date, origin = ymd("1899-12-30"))
  )
# Join demand, temperatures and holidays
aus_elec <- demands %>%
  left_join(temperatures, by = c("State", "Date", "Period")) %>%
  transmute(
    State,
    Time = as.POSIXct(Date + minutes((Period-1) * 30)),
    Period,
    Date = as_date(Time),
    DOW = wday(Date, label=TRUE),
    Demand = OperationalLessIndustrial, 
    Temperature = Temp,
  ) %>%
  left_join(holidays, by = c("State", "Date")) %>%
  replace_na(list(Holiday = FALSE))
# Remove duplicates and create a tsibble
aus_elec <- aus_elec %>%
  filter(!are_duplicated(aus_elec, index=Time, key=State)) %>%
  as_tsibble(index = Time, key=State)
```

This block of code reads in raw data files containing holiday information, temperatures and electricity demand for each state, and then joins them into a single tsibble. For some reason, there are duplicated rows from South Australia, so the last few lines removes the duplicates before forming a tsibble, keyed by State.

```{r ausdata2, dependson='ausdata'}
aus_elec
```

This data set contains half-hourly data from all states from 1 January 2002 - 1 March 2015 (and in the case of Queensland to 1 April 2015). The temperature variable is from a weather station in the capital city of each state.

&nbsp;

## GEFCOM 2017

The [Global Energy Forecasting Competition in 2017](http://www.drhongtao.com/gefcom/2017) involved data on hourly zonal loads of ISO New England from March 2003 to April 2017. The data have already been packaged into tibble format by Cameron Roach in the [gefcom2017data Github repository](https://github.com/camroach87/gefcom2017data). So it is relatively easy to convert this to a tsibble.

```{r gefcom2017, message=FALSE, warning=FALSE}
devtools::install_github("camroach87/gefcom2017data")
library(gefcom2017data)
gefcom2017 <- gefcom %>% 
  ungroup() %>%
  as_tsibble(key=zone, index=ts)
gefcom2017
```

Details of the data (and the competition) are available on [Tao Hong's website](http://blog.drhongtao.com/2016/10/gefcom2017-hierarchical-probabilistic-load-forecasting.html).