---
author: robjhyndman
comments: true
date: 2022-06-05
title: "The cricketdata package"
slug: cricketdata
output:
  html_document:
    keep_md: true
mathjax: true
categories:
  - R
  - data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  cache = TRUE
)
options(digits = 3, width = 75)

# Okabi-Ito colours
options(
  ggplot2.discrete.colour = c("#D55E00", "#0072B2","#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442")
)
```

## Four functions

[The cricketdata package](http://pkg.robjhyndman.com/cricketdata/) has been around for a few years [on github](https://github.com/robjhyndman/cricketdata), and it has been on [CRAN](https://cloud.r-project.org/package=cricketdata) since February 2022. There are only four functions:

 * `fetch_cricinfo()`: Fetch team data on international cricket matches provided by [ESPNCricinfo](https://www.espncricinfo.com/).
 * `fetch_player_data()`: Fetch individual player data on international cricket matches provided by [ESPNCricinfo](https://www.espncricinfo.com/).
 * `find_player_id()`: Search for the player ID on [ESPNCricinfo](https://www.espncricinfo.com/).
 * `fetch_cricsheet()`: Fetch ball-by-ball, match and player data from [Cricsheet](https://cricsheet.org).

[Jacquie Tran](https://www.jacquietran.com) wrote the first version of the `fetch_cricsheet()` function, and [the vignette](http://pkg.robjhyndman.com/cricketdata/articles/cricsheet.html) which demonstrates it.

Here are some examples demonstrating the Cricinfo functions.

```{r packages, message=FALSE}
library(cricketdata)
library(tidyverse)
```

## Women's T20 bowling data

The `fetch_cricinfo()` function downloads data for international T20, ODI or Test matches, for men or women, and for batting, bowling or fielding. By default, it downloads career-level statistics for individual players. Here is an example for women T20 bowlers.

```{r wt20}
# Fetch all Women's T20 data
wt20 <- fetch_cricinfo("T20", "Women", "Bowling")
```

```{r woment20}
wt20 %>%
  select(Player, Country, Matches, Runs, Wickets, Economy, StrikeRate)
```

We can plot a bowler's strike rate (balls per wicket) vs economy rate (runs per wicket). Each observation represents one player, who has taken at least 50 international wickets.

```{r woment20graph}
wt20 %>%
  filter(Wickets >= 50) %>%
  ggplot(aes(y = StrikeRate, x = Average)) +
  geom_point(alpha = 0.3, col = "blue") +
  ggtitle("Women International T20 Bowlers") +
  ylab("Balls per wicket") + xlab("Runs per wicket")
```

```{r boochatham, echo=FALSE}
boochatham <- wt20 %>%
  filter(Player == "N Boochatham")
```

The extraordinary result on the bottom left is due to the Thai all-rounder, Nattaya Boochatham, who has taken `r boochatham$Wickets` wickets, with a strike rate of `r boochatham$StrikeRate`, an average of `r boochatham$Average`, and an economy rate of `r boochatham$Economy`.

## Australian men's ODI data by innings

The next example shows Australian men's ODI batting results by innings.

```{r menodidownlaod}
# Fetch all Australian Men's ODI data by innings
menODI <- fetch_cricinfo("ODI", "Men", "Batting", type = "innings", country = "Australia")
```

```{r menodi}
menODI %>%
  select(Date, Player, Runs, StrikeRate, NotOut)
```

```{r menodiave, echo=FALSE}
ave <- menODI %>%
  filter(
    Date >= "2000-01-01",
    !is.na(Runs) & !is.na(NotOut)
  ) %>%
  select(Runs, NotOut)

ave <- sum(ave$Runs)/sum(!ave$NotOut)
```

```{r menodigraph, warning=FALSE, message=FALSE}
menODI %>%
  ggplot(aes(y = Runs, x = Date)) +
  geom_point(alpha = 0.2, col = "#D55E00") +
  geom_smooth() +
  ggtitle("Australia Men ODI: Runs per Innings")
```

The average number of runs per innings slowly increased until about 2000, after which it has remained largely constant at about `r round(ave,1)`. This is a little higher than the smooth line shown on the plot, which has not taken account of not-out results.

## Indian test fielding data

Next, we demonstrate some of the fielding data available, using Test match fielding from Indian men's players.

```{r indiafieldingdownload}
Indfielding <- fetch_cricinfo("Test", "Men", "Fielding", country = "India")
```

```{r indiafielding}
Indfielding
```

We can plot the number of dismissals by number of matches for all male test players. Because wicket keepers typically have a lot more dismissals than other players, they are shown in a different colour.

```{r indiafieldinggraph}
Indfielding %>%
  mutate(wktkeeper = (CaughtBehind > 0) | (Stumped > 0)) %>%
  ggplot(aes(x = Matches, y = Dismissals, col = wktkeeper)) +
  geom_point() +
  ggtitle("Indian Men Test Fielding")
```

The high number of dismissals, close to 300, is of course due to MS Dhoni. Another interesting one here is the non-wicketkeeper with over 200 dismissals, which is Rahul Dravid who took 209 catches during his career.

## Meg Lanning's ODI batting

Finally, let's look at individual player data. The `fetch_player_data()` requires the Cricinfo player ID, which you can either look up on their website, or use the `find_player_id()` function. We will look at the ODI results of Australia's captain, Meg Lanning.

```{r meglanningsetup}
meg_lanning_id <- find_player_id("Lanning")$ID
MegLanning <- fetch_player_data(meg_lanning_id, "ODI") %>%
  mutate(NotOut = (Dismissal == "not out"))
```

```{r meglanning}
MegLanning
```

We can plot her runs per innings on the vertical axis over time on the horizontal axis.

```{r meglanninggraph}
# Compute batting average
MLave <- MegLanning %>%
  filter(!is.na(Runs)) %>%
  summarise(Average = sum(Runs) / (n() - sum(NotOut))) %>%
  pull(Average)
names(MLave) <- paste("Average =", round(MLave, 2))
# Plot ODI scores
ggplot(MegLanning) +
  geom_hline(aes(yintercept = MLave), col="gray") +
  geom_point(aes(x = Date, y = Runs, col = NotOut)) +
  ggtitle("Meg Lanning ODI Scores") +
  scale_y_continuous(sec.axis = sec_axis(~., breaks = MLave))
```

She has shown amazing consistency over her career, with centuries scored in every year of her career except for 2021, when her highest score from 6 matches was 53.
